A note on the instability and degeneracy of deep learning models

### 2017-11-06 writing/paper-note/docs/note_blinded.tex
ready?

\maketitle
\begin{abstract}
A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. For correlated data structures found in several application
areas, there is increasing interest in predicting/identifying such
sensitivity in model probability structure. We consider the problem of
quantifying instability for general probability models defined on
sequences of observations, where each sequence of length \(N\) has a
finite number of possible outcomes. A sequence of probability models
results, indexed by \(N\), that accommodates data of expanding
dimension. Model instability is formally shown to occur when a certain
log-probability ratio under such models grows faster than \(N\). In this
case, a one component change in the data sequence can shift probability
by orders of magnitude. Also, as instability becomes more extreme, the
resulting probability models are shown to tend to degeneracy, placing
all their probability on potentially small portions of the sample space.
These results on instability apply to large classes of models commonly
used in random graphs, network analysis, and machine learning contexts.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
parameters can be tuned in sign with effects prescribed in the model
condition PSR, unstable FOES models will exhibit further probability
sensitivities, as outlined in the following extension of Theorem
@ref(thm:degenFOES\}).

@@

\BeginKnitrBlock{remark}
\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
\ref{cor:sign} can also be extended to cases where parameter components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
more generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
involving a switch in sign only among some dominating model parameters
\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
which dominate the probability structure of the model, then the results
of Corollary \ref{cor:sign} can still hold with
\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
example of one sufficient condition, if
\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
holds in addition to Corollary\textasciitilde{}1 assumptions, where \[
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
\] represents a standardized form of \(\boldsymbol \theta\)-model
probabilities, then the results of Corollary \ref{cor:sign} apply to
\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
then imply that many more unstable models exist over a broader spectrum
of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
\(\boldsymbol \theta_N\), which involves some amount of sign change
among components of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

\protect\hypertarget{prp:prop2}{}{\label{prp:prop2}}Let

@@

    based on $f_{\boldsymbol \theta_N}$ from \\eqref{eq:f} with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.

@@

visible model as well).

@@

Volatility in the probability structure of an unstable model can also
hamper efforts to maximize likelihood functions in statistical
inference. When a FOES model is unstable along a parameter sequence
\(\boldsymbol \theta_N\), the same model can further be unstable along
parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
origin (model condition PSR and Corollary \ref{cor:sign}). This can
translate into potential sensitivity of the likelihood function around
zero, and lead to numerical complications in maximizing the objective
function. We next provide a discussion of this issue in a way that
builds upon and extends related findings by @Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
largely focused on the case of one-parameter exponential models.

>>>>

\maketitle
\begin{abstract}
A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. This instability is a property of the probability model,
rather than the fitted parameter vector. For correlated data structures
found in several application areas, there is increasing interest in
predicting/identifying such sensitivity in model probability structure.
We consider the problem of quantifying instability for general
probability models defined on sequences of observations, where each
sequence of length \(N\) has a finite number of possible values. A
sequence of probability models results, indexed by \(N\), that
accommodates data of expanding dimension. Model instability is formally
shown to occur when a certain log-probability ratio under such models
grows faster than \(N\). In this case, a one component change in the
data sequence can shift probability by orders of magnitude. Also, as
instability becomes more extreme, the resulting probability models are
shown to tend to degeneracy, placing all their probability on
potentially small portions of the sample space. These results on
instability apply to large classes of models commonly used in random
graphs, network analysis, and machine learning contexts.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
parameters can be tuned in sign with effects prescribed in the model
condition PSR, unstable FOES models will exhibit further probability
sensitivities, as outlined in the following extension of Theorem
\ref{thm:degenFOES}.

@@

\BeginKnitrBlock{remark}
\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
\ref{cor:sign} can also be extended to cases where parameter components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
more generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
involving a switch in sign only among some dominating model parameters
\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
which dominate the probability structure of the model, then the results
of Corollary \ref{cor:sign} can still hold with
\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
example of one sufficient condition, if
\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
holds in addition to Corollary \ref{cor:sign} assumptions, where \[
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
\] represents a standardized form of \(\boldsymbol \theta\)-model
probabilities, then the results of Corollary \ref{cor:sign} apply to
\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
then imply that many more unstable models exist over a broader spectrum
of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
\(\boldsymbol \theta_N\), which involves some amount of sign change
among components of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let

@@

    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.

@@

visible model as well). Further evidence of this is seen in the
following numerical example. \par
\begin{figure}
\includegraphics{note_files/figure-latex/rbm-plots-1} \includegraphics{note_files/figure-latex/rbm-plots-2} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/N_{\mathcal{V}}$ (left) and $\DN(\boldsymbol \theta)$ at each grid point for each combination of magnitude of $\boldsymbol \theta$. As the magnitude of $\boldsymbol \theta$ grows, so does the value of these metrics, indicating typical instability in the model.}\label{fig:rbm-plots}
\end{figure}
In our numerical experiment, we allow the two types of terms (main
effects terms corresponding to visible and hidden parameters
\(\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})\)
and interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\)) to
have varying average magnitudes,
\(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\)
\(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\)
for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\)
hiddens. These average magnitudes vary on a grid between 0.001 and 3
with 20 breaks, yielding 400 grid points. At each point in the grid, 100
vectors (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a
sphere with radius corresponding to the first coordinate in the grid and
100 vectors (\(\boldsymbol \theta_{interaction}\)) are sampled uniformly
on a sphere with radius corresponding to the second coordinate in the
grid via sums of squared and scaled iid Normal\((0, 1)\) variables.
These vectors are then paired to create 100 values of
\(\boldsymbol \theta_N\) with magnitudes at each point in the grid. The
values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\) and
\(\DN(\boldsymbol \theta_N)\) are then calculated for each
\(\boldsymbol \theta_N\) and then summarized for each point in the grid
using the sample mean. The results of this numerical study are shown in
Figure \ref{fig:rbm-plots}. From these two plots, it is clear that for
larger magnitudes of the parameter vectors, there is evidence of
S-instability in that the log-ratio of extremal probabilities scaled by
\(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one
component change in data outcomes are both increasing away from
\(\boldsymbol \theta_N = \boldsymbol 0\), further supporting
\ref{prp:prop2}(iii.2 and iii.5).

@@

Volatility in the probability structure of an unstable model can also
hamper efforts to maximize likelihood functions in statistical
inference. When a FOES model is unstable along a parameter sequence
\(\boldsymbol \theta_N\), the same model can further be unstable along
parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
origin (model condition PSR and Corollary \ref{cor:sign}). This can
translate into potential sensitivity of the likelihood function around
zero, and lead to numerical complications in maximizing the objective
function. We next provide a discussion of this issue in a way that
builds upon and extends related findings by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
largely focused on the case of one-parameter exponential models.

### 2017-11-06 writing/paper-note/index.Rmd
dr. nordman's changes
ready?

---
title: "A note on the instability and degeneracy of deep learning models"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. A sequence of probability models results, indexed by $N$, that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

@@

library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(mvtnorm)
library(rootSolve)
library(agridat)
library(xtable)
```

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. This instability is a property of the probability model, rather than the fitted parameter vector. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible values. A sequence of probability models results, indexed by $N$, that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

@@

library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(scales)
library(xtable)
```
